Joint maximum-likelihood of phylogeny and ancestral states is not consistent

### 2019-03-21 versions/response.v2.txt

Thank you for handling the manuscript, and please thank the reviewers for their helpful comments. As stated by Reviewer 3, we believe that the review process has significantly improved the manuscript.

> Reviewers' comments:
> Reviewer: 1

> Comments to the Author
> As with the previous version, I believe the topic matter will be of
> interest. For me, however, the work is incomplete in important ways
> that I believe can be relatively easily addressed. The authors went
> beyond my expectations in calculating explicit solutions for the
> limiting edge-lengths (Fig S2) when tau1 is the putative topology of
> estimation. This allows them to draw conclusions about edge-length
> inconsistency but there is no definitive conclusion about `strict'
> topological inconsistency (in the sense of converging upon the wrong
> tree rather than a multifurcating one). From the statements near l238
> it seems the authors have completed some preliminary investigation on
> strict topological inconsistency and find no evidence for it at
> present. Assuming that remains true, the findings of the current ms
> are of a different nature than the previous version where strict
> topological inconsistency was found. In settings with a limiting
> multifurcating topology, because it is multifurcating, for finite but
> large sequence-lengths, measures of uncertainty (eg. bootstrap
> support) will suggest that there is substantial uncertainty in
> whatever topology gets estimated (this probably deserves discussion).
> The joint optimization approach thus lacks the power to estimate the
> correct topology where ML would but is not being positively mislead.
> Assuming the lack of strict topological inconsistency holds, the
> results for the method are not as negative as they were in the
> previous version.

DISCUSS I honestly can't figure out what the reviewer wants to see re the bootstrap. Branch length inferences don't come into the bootstrap unless you contract things with zero branch lengths, which isn't generally done when doing bootstrap estimation.

@@

The reviewer is correct that this story is different than the one in the previous version of the paper.

DISCUSS Shall we strike the speculation about higher likelihoods on the wrong topology?

@@

I'm afraid that this fact wasn't obvious to us at the outset of this project. The interest that the other reviewers have shown in the manuscript confirms that it is not broadly obvious.

> 2) In some parameter regions the internal branch of the four taxon
> tree collapses to 0.
> Again, this is true and identifying regions in parameter space where
> the internal branch collapses to zero is the most interesting part of
> the manuscript in my opinion. Unfortunately, the manuscript doesn't
> give much insight as to why the length of the internal branch shrinks
> to zero in some regions of parameter space. There is, however, a
> rather intuitive explanation for this. In the inv Fels topology, the
> internal branch collapses to 0 if
>
> p_x^2 > c p_y
>
> (the prefactor c depends on specifics like alphabet size etc). And
> this has an intuitive interpretation: if p_x^2 > p_y, it is more
> likely to observe a mutation twice on two branches of length p_x than
> once on a branch of length p_y. As a result, the joint inference will
> push p_w to zero and assign all state changes on the terminal branches
> of length p_x. But for short branches, this corresponding sector of
> parameter space is a small (tending to zero for a fixed p_x/p_y with
> p_x \to 0).
>
> My empirical tests of the approximate joint inference scheme suggest
> this region is much smaller than the one identified by Shaw et al in
> Fig 3.

We have worked this out with Dr. Neher and found that the difference lay in that he was not optimizing the objective function from both sides. This simulation work also demonstrated that the intuitive explanation by this reviewer is incorrect.

@@

> Perhaps it's a latexdiff leftover but eq 6 in line 151 seems to have a
> line with just a ; \xi )

Indeed, this was not present in the non-latexdiff pdf.

> Line 593 p 25 I don't understand "the mapping \psi operates by
> returning the tips labeled as 1 in a site pattern to obtain a site
> split in P(Y) if the set of tips labeled 1 is not in P(Y)" and this
> mapping seems fundamental. Perhaps you mean to refer to the m'th taxon
> (or node)..?

We have rewritten this confusing sentence in terms of complements, and have added a little bit of text pointing the reader to the formal definition just above.

@@

> The symmetric model is assumed strongly in the first part of the
> appendix, with the factor of 2; my guess is that the results hold in
> asymmetric models but the notation is harder, is that correct?

Unfortunately a lot would have to change to generalize from the symmetric model. We would not be able to "collapse" site splits, and the Hadamard transform only works for submodels of the Kimura 3 parameter model. See, for example:

Bryant, D. (2009). Hadamard Phylogenetic Methods and the n-taxon Process. Bulletin of Mathematical Biology, 71(2), 339–351. https://doi.org/10.1007/s11538-008-9364-8

@@

> What's q in equation 7 in the appendix (line 655) ?

We now remind the reader that this is the number of site splits.

@@

> More substantitvely, the limit in (8) seems not established because as
> n -> \infty so would q and n_j, so these can't be treated as fixed .

We now emphasize that the number of taxa is fixed so q is fixed, and that although n_j goes to infinity the ratio n_j/n converges to the probability of seeing that column.

@@

> page 9 - at this point in the paper, unless the reader has already
> read the appendix they don't know what the estimator means or what the
> p_ytilde j are - so this text doesn't make sense at this point.

This is a good point! We have rephrased things, eliminating p_ytilde_j except for a parenthetical reference to the Appendix. We believe that the result is much more readable.

@@

> page 30 line 667-8 what are S and Y and what makes a path? I gather
> from the math that you mean paths in the tree, since the sum is over
> ancestral states and there's a product of probabilities of changes on
> branches but (perhaps my lack of background) the form of (9) isn't
> obvious I think because I am not exactly clear on what the notation
> means.

We have clarified the definitions of S and Y, and added that by path we mean graph-theoretic path. We are afraid that adding more detail about the Hadamard transform is beyond the scope of this paper, though we have added a link to a scanned page of the Semple and Steel book where I (Erick) have added some details on the proofs.

### 2019-03-21 appendix.tex
Done with everything I can do.

We first fix notation.
\begin{definition}
Let the mapping from site patterns to site splits
\[
\patternToSplit:\alphabet^\nSiteRows\rightarrow\mathcal{P}(\siteSplitSet)
\]
be
\[
\patternToSplit(\alignmentColumn) =
\left\{
    \begin{array}{ll}
        \{i'\in\{1,\ldots,\nSiteRows-1\}: \alignmentColumn_{i,i'}=1\}  & \mbox{if } \alignmentColumn_{i,\nSiteRows}=0,\\
        \{i'\in\{1,\ldots,\nSiteRows-1\}: \overline{\alignmentColumn}_{i,i'}=1\}  & \mbox{if } \alignmentColumn_{i,\nSiteRows}=1,
    \end{array}
\right.
\]
and the mapping from ancestral states and tip states to ancestral state splits
\[
\ancestralToSplit:\alphabet^\nSiteRows\times\alphabet^\nAncestralStateRows\rightarrow\mathcal{P}(\ancestralSplitSet)
\]
be
\[
\ancestralToSplit(\alignmentColumn, \ancestralStateColumn) =
\left\{
    \begin{array}{ll}
        \{i'\in\{1,\ldots,\nAncestralStateRows\}: \ancestralStateColumn_{i,i'}=1\}  & \mbox{if } \alignmentColumn_{i,\nSiteRows}=0,\\
        \{i'\in\{1,\ldots,\nAncestralStateRows\}: \overline{\ancestralStateColumn}_{i,i'}=1\}  & \mbox{if } \alignmentColumn_{i,\nSiteRows}=1.
    \end{array}
\right.
\]
Then, given a site pattern--valued random variable $\alignmentColumnRV$ and an ancestral state--valued random variable $\ancestralStateColumnRV$, define the random variables
\[
\siteSplitRV := \patternToSplit(\alignmentColumnRV)
\]
and
\[
\ancestralSplitRV := \ancestralToSplit(\alignmentColumnRV, \ancestralStateColumnRV).
\]
\end{definition}
The mapping $\patternToSplit$ operates by returning the tips labeled as $1$ in a site pattern to obtain a site split in $\mathcal{P}(\siteSplitSet)$ if the set of tips labeled $1$ is not in $\mathcal{P}(\siteSplitSet)$.
The mapping $\ancestralToSplit$ is defined by whether the tip states have their complements taken or not: if the set of tips labeled $1$ in $\alignmentColumn$ is in $\mathcal{P}(\siteSplitSet)$, $\ancestralToSplit(\alignmentColumn, \ancestralStateColumn)$ is the set of tips labeled $1$ in $\ancestralStateColumn$; otherwise, the set of tips labeled $1$ in $\overline{\alignmentColumn}$ necessarily is in $\mathcal{P}(\siteSplitSet)$ and so $\ancestralToSplit(\alignmentColumn, \ancestralStateColumn)$ is $\overline{\ancestralStateColumn}$.

@@

Let
\[
L_\nCols''(\tau, t; \fullAlignment) = \prod_{j=1}^{\nSiteSplits} \ \left[\Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_{\siteSplit_j} \mid \tau, t)\right] ^{\nCols_j(\fullAlignment)}
\]
be the final product in \eqref{eq:site_pattern_likelihood}.
Assume $\nCols$ observations are generated from a model with parameters $(\tau^*, t^*)$.
We have
\begin{equation*}
\frac{1}{\nCols} \log L_\nCols''(\tau, t; \fullAlignment) = \sum_{j=1}^\nSiteSplits \frac{\nCols_j(\fullAlignment)}{\nCols}\cdot  \log \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_{\siteSplit_j} \mid \tau, t)
\end{equation*}
so that, in the $\nCols\rightarrow\infty$ limit,
\begin{equation}
\begin{split}
&    \frac{1}{\nCols} \log L_\nCols''(\tau, t; \fullAlignment) \\
&\qquad \rightarrow \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot \log \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_{\siteSplit_j} \mid \tau, t). \label{eq:site_pattern_profile_likelihood_mean}
\end{split}
\end{equation}
%EM das I believe this is the definition of \ell-- this should be made more explicit.
%EM Also, in the main text we read that \ell is the limit of L' as n goes to infinity, but here it's L''.

@@

We state the Hadamard representation of site split generating probabilities---that is, probabilities of obtaining particular site splits given a tree---following Section 8.6 of \citet{Semple2003-em}.
For each edge $e$ define the edge ``fidelity'' for that edge as
\[
\theta(e) = 1-2p(e)
\]
where $p(e)$ is the probability of a character change along edge $e$.
For an even-sized subset of $Y\subseteq\mathcal{S}$, let the path set $P(Y)$ be the set of edges in the path connecting both elements of $Y$.
For $n$ taxa, the probability of observing site split $A\in\mathcal{P}(\siteSplitSet)$ is
\begin{equation}
\label{eq:hadamard_probability}
p_A = \frac{1}{2^{n-1}} \ \sum_{Y \subseteq \mathcal{S} : |Y| \equiv 0 (\mathrm{mod} \ 2)} \ \left[(-1)^{|Y \cap A|} \ \prod_{e\in P(Y)} \ \theta(e) \right].
\end{equation}
By convention, we set $P(\emptyset)=\emptyset$ and $\prod_{e\in\emptyset} \ \theta(e) = 1$.
For notational convenience, let
\[
p_{\siteSplit_j} := \Pr(\siteSplitRV=\siteSplit_j \mid \tau_1,t),
\]
for any site split $\siteSplit_j$.
Table~\ref{tab:gen-sitepatprob} contains calculations of site split probabilities for the trees in Fig.~\ref{fig:farris-fels-top}.

>>>>

We first fix notation.
\begin{definition}
Let the mapping from site patterns to site splits
\[
\patternToSplit:\alphabet^\nSiteRows\rightarrow\mathcal{P}(\siteSplitSet)
\]
be
\[
\patternToSplit(\alignmentColumn) =
\left\{
    \begin{array}{ll}
        \{i'\in\{1,\ldots,\nSiteRows-1\}: \alignmentColumn_{i,i'}=1\}  & \mbox{if } \alignmentColumn_{i,\nSiteRows}=0,\\
        \{i'\in\{1,\ldots,\nSiteRows-1\}: \overline{\alignmentColumn}_{i,i'}=1\}  & \mbox{if } \alignmentColumn_{i,\nSiteRows}=1,
    \end{array}
\right.
\]
and the mapping from ancestral states and tip states to ancestral state splits
\[
\ancestralToSplit:\alphabet^\nSiteRows\times\alphabet^\nAncestralStateRows\rightarrow\mathcal{P}(\ancestralSplitSet)
\]
be
\[
\ancestralToSplit(\alignmentColumn, \ancestralStateColumn) =
\left\{
    \begin{array}{ll}
        \{i'\in\{1,\ldots,\nAncestralStateRows\}: \ancestralStateColumn_{i,i'}=1\}  & \mbox{if } \alignmentColumn_{i,\nSiteRows}=0,\\
        \{i'\in\{1,\ldots,\nAncestralStateRows\}: \overline{\ancestralStateColumn}_{i,i'}=1\}  & \mbox{if } \alignmentColumn_{i,\nSiteRows}=1.
    \end{array}
\right.
\]
Then, given a site pattern--valued random variable $\alignmentColumnRV$ and an ancestral state--valued random variable $\ancestralStateColumnRV$, define the random variables
\[
\siteSplitRV := \patternToSplit(\alignmentColumnRV)
\]
and
\[
\ancestralSplitRV := \ancestralToSplit(\alignmentColumnRV, \ancestralStateColumnRV).
\]
\end{definition}
To explain the above definition in words, the mapping $\patternToSplit$ operates by taking the complement of a site pattern to obtain a site split in $\mathcal{P}(\siteSplitSet)$ if the set of tips labeled $1$ is not in $\mathcal{P}(\siteSplitSet)$.
The mapping $\ancestralToSplit$ is defined by whether the tip states have their complements taken or not: if the set of tips labeled $1$ in $\alignmentColumn$ is in $\mathcal{P}(\siteSplitSet)$, $\ancestralToSplit(\alignmentColumn, \ancestralStateColumn)$ is the set of tips labeled $1$ in $\ancestralStateColumn$; otherwise, the set of tips labeled $1$ in $\overline{\alignmentColumn}$ necessarily is in $\mathcal{P}(\siteSplitSet)$ and so $\ancestralToSplit(\alignmentColumn, \ancestralStateColumn)$ is $\overline{\ancestralStateColumn}$.

@@

Let
\[
L_\nCols''(\tau, t; \fullAlignment) = \prod_{j=1}^{\nSiteSplits} \ \left[\Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_{\siteSplit_j} \mid \tau, t)\right] ^{\nCols_j(\fullAlignment)}
\]
be the final product in \eqref{eq:site_pattern_likelihood}.
Assume $\nCols$ observations are generated from a model with parameters $(\tau^*, t^*)$.
We have
\begin{equation*}
\frac{1}{\nCols} \log L_\nCols''(\tau, t; \fullAlignment) = \sum_{j=1}^\nSiteSplits \frac{\nCols_j(\fullAlignment)}{\nCols}\cdot  \log \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_{\siteSplit_j} \mid \tau, t)
\end{equation*}
so that, in the $\nCols\rightarrow\infty$ limit,
\begin{equation}
\begin{split}
&    \frac{1}{\nCols} \log L_\nCols''(\tau, t; \fullAlignment) \\
&\qquad \rightarrow \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot \log \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_{\siteSplit_j} \mid \tau, t) \label{eq:site_pattern_profile_likelihood_mean}
\end{split}
\end{equation}
because
$\frac{\nCols_j(\fullAlignment)}{\nCols}$ converges to $\Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)$.
%EM das I believe this is the definition of \ell-- this should be made more explicit.
%EM Also, in the main text we read that \ell is the limit of L' as n goes to infinity, but here it's L''.
Here $j$ indexes the site splits, of which there are a fixed number $q$.

@@

We state the Hadamard representation of site split generating probabilities---that is, probabilities of obtaining particular site splits given a tree---following Section 8.6 of \citet{Semple2003-em}.
For additional detail on the proofs in that section, see \url{http://i.imgur.com/T5XLPLB.jpg}.
For each edge $e$ define the edge ``fidelity'' for that edge as
\[
\theta(e) = 1-2p(e)
\]
where $p(e)$ is the probability of a character change along edge $e$.
For an even-sized subset $Y$ of a taxon set $\mathcal{S}$, let the path set $P(Y)$ be the set of edges in the graph-theoretic path connecting both elements of $Y$.
For $n$ taxa, the probability of observing site split $A\in\mathcal{P}(\siteSplitSet)$ is
\begin{equation}
\label{eq:hadamard_probability}
p_A = \frac{1}{2^{n-1}} \ \sum_{Y \subseteq \mathcal{S} : |Y| \equiv 0 (\mathrm{mod} \ 2)} \ \left[(-1)^{|Y \cap A|} \ \prod_{e\in P(Y)} \ \theta(e) \right].
\end{equation}
By convention, we set $P(\emptyset)=\emptyset$ and $\prod_{e\in\emptyset} \ \theta(e) = 1$.
For notational convenience, let
\[
p_{\siteSplit_j} := \Pr(\siteSplitRV=\siteSplit_j \mid \tau_1,t),
\]
for any site split $\siteSplit_j$.
Table~\ref{tab:gen-sitepatprob} contains calculations of site split probabilities for the trees in Fig.~\ref{fig:farris-fels-top}.

### 2019-03-21 joint_inf.tex
Done with everything I can do.

In words, the joint estimation procedure does not recover the true generating $t^*$ almost everywhere in the space of generating parameters.
The proof that $\hat{t} \neq t^*$ everywhere except a set of measure zero can be seen intuitively through the fact that the estimator for a given branch fidelity cannot isolate the individual generating parameter as a linear term since the estimator itself is a combination of nonlinear functions in $(x^*,y^*)$ of generating fidelities.
For example, in estimating $x_1$, $\hat{x}_1$ is a linear combination of $p_{\siteSplit_j}$ values.
For the generating probabilities (top panel of Table~\ref{tab:gen-sitepatprob}), there is no linear combination that results in an isolated $x^*$ term, as all terms are either quadratic, i.e., $(x^*)^2$ or $(x^*)^2(y^*)^2$, or have both $x^*$ and $y^*$, i.e., $x^*(y^*)^2$.
Thus, we cannot obtain a linear combination that in the general case yields $x^*$.
We may have special cases where certain $x^*$ and $y^*$ values yield consistent estimates for $\hat{t}$, but consistency does not hold in general.
We give an example of consistency in a degenerate case in the Appendix following the proof of Theorem~\ref{thm:bl_theorem}.

>>>>

In words, the joint estimation procedure does not recover the true generating $t^*$ almost everywhere in the space of generating parameters.
The fact that $\hat{t} \neq t^*$ everywhere except a set of measure zero can be seen intuitively through the fact that the estimator for a given branch fidelity cannot isolate the individual generating parameter as a linear term since the estimator itself is a combination of nonlinear functions in $(x^*,y^*)$ of generating fidelities.
For example, in estimating $x_1$, $\hat{x}_1$ is a linear combination of data-generating probabilities; these probabilities are polynomials in $x^*$ and $y^*$.
For the generating probabilities (these are called $p_{\siteSplit_j}$ values in the Appendix; see Table~\ref{tab:gen-sitepatprob} for a complete tabulation), there is no linear combination that results in an isolated $x^*$ term, as all terms are either quadratic, i.e., $(x^*)^2$ or $(x^*)^2(y^*)^2$, or have both $x^*$ and $y^*$, i.e., $x^*(y^*)^2$.
Thus, we cannot obtain a linear combination that in the general case yields $x^*$.
Although there may be special cases in which certain $x^*$ and $y^*$ values yield consistent estimates for $\hat{t}$, but consistency does not hold in general.
We give an example of consistency in a degenerate case in the Appendix following the proof of Theorem~\ref{thm:bl_theorem}.
